Towards a Wide-Coverage Grammar for Swedish Using GF

### 2012-01-07 doc/report/report.tex
small bugs fixed all over

\abstract{
The thesis describes the work to ~~achieve a covering Swedish~~ grammar implemented
%TODO first sentence should be more, no understatements!
in GF -- to model natural language by functional programming. 
The idea has been to combine existing language technology resources with
new techniques, and the aim to obtain a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
\\We present the methods and technologies used and further inspects the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and a Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
extends it to cover constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed and our
goal is
to create an open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
%TODO verify this
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules describing morphology and syntax.
From now on, this is what we mean by grammar, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get %and thereby get
%The usage of GF allows us to start from
a well-defined system for describing the language, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

\newpage % until it looks nicer
\section{Outline}
The thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction to the Swedish language, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med")) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (``den" and ``honom").
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{`You look at the boy.'}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=80mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=80mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}

@@

In addition to the core resource grammars, which is shared with the other languages implemented in the
library, 
%In addition to the basic resource grammar,
 language specific constructions may bee added
%>> core is shared but particular lang. has particular features.
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation: %$ although not all
%of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
%For example, \verb|FocObj| fronted the object as in sentence (\ref{gfSwe:apple}).
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &have}
{`I don't want that apple'}}\label{gfSwe:apple}
and for preposition stranding:
\enumsentence{
~~Stranded preposition \hspace{45mm} cf.\\
\begin{tabular}{ll}
{\begin{tabular}{llllll}
Vem& måste& jag& akta& mig &\textbf{för}? \\
\emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me} &\textbf{\emph{for}}? \\
\end{tabular}}&
{\begin{tabular}{llllll}
\textbf{För}& vem& måste& jag& akta& mig?\\
\emph{\textbf}{for}& \emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me}?\\
\end{tabular}}\\
~~`Who do I need to watch out \textbf{for}?' &~ `\textbf{For} whom do I need to watch out?'\\
\end{tabular}}

@@

\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw {\sc self's} children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
{\sc self's} gloves, he had forgotten on the train.}
%\enumsentence{Han ger sina pengar till sina syskon.\\
%He gives {\sc self's} money to {\sc self's} children.}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all {\sc self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
a. & Han är längre än sin kompis. & & b. &Han är här oftare än sin kompis.\\
   & He is taller than {\sc self's} friend.
   & & & He is here more often than {\sc self's} friend.
\end{tabular}}\label{ex:ReflAP} \label{ex:ReflAdv}
\eenumsentence{%\begin{tabular}[t]{@{}*{5}{l@{\ }}}
\item Hon tyckte om skolan och alla sina elever.\\
     She liked the school and all {\sc self's} students.
\item Han såg sina få böcker och sin penna.\\
 He saw {\sc self's} few books and {\sc self's} pencil.
}\label{ex:ReflConj} \label{ex:ReflConj2}
Reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in %, it is hence not
%allowed as subject. The antecedent must not be out of scope for the closest finite verb,
sentence (\ref{sent:sesig}) and (\ref{ex:ReflSubj}).
The third person reflexives (`sig',`sin') requires a third person antecedent (see \ref{ex:refljag}).
Furthermore, the antecedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).
The grammar should not accept any of these sentences:
\enumsentence{
*Sina vantar var kvar på tåget.\\
{\sc self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Hon och sin kompis läser en bok.\\
She and {\sc self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give {\sc self's} money to {\sc self's} siblings. }\label{ex:refljag}
\enumsentence{*Han vill att jag ser på sig.\\
He wants me to look at {\sc self}.}\label{sent:sesig}

@@

Apart for this, only corrections have been done, exemplified in
this sentence.
\enumsentence{
Hon sover \textbf{som} är bra ~ \textrightarrow ~ Hon sover \textbf{vilket} är bra\\
``She sleeps, which is good"}

@@

\chapter{Discussion}
\label{sec:results}
\section{Evaluation}
%TODO
%>>evolution diagram?
%also removed unnecessary ambigutiies.
%>> how to handle ambigs: know them and controll them. using statistics or ontolgies.
%>> that is: disambiguitation by context or by meaning.
The project has resulted in
\begin{itemize}
\item a large-scale GF lexicon and a program to redo the importation when needed
\item a comparison between GF and another another annotation
\item a deeper testing of the Swedish resource grammar and an estimation
of how well GF can be used to describe larger parts of a language
\item a study of how dependent types can be used in the resource grammars
\end{itemize}

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which is fast even with an extensive lexicon.
%TODO make nice!
Although it cannot not been verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{I'm getting hungry}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{I'm getting hungry}}\label{sent:complVV1}
Since there is no information about which verbs that allows the infinitive mark
to be left out, it will also allow the more questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & vill & \textbf{att}& du & går} 
{I & want & \textbf{to} &you & leave}
{I want you to leave}
\item \shortexnt{4}
{Jag & vill & du & går} 
{I & want &you & leave}}\label{sent:complVV2}
%ComplBareVV, SupCl
For random generation, the sentences generated may be next to impossible
to interpret without looking at the abstract tree, since they may contain
too many attributes, relative clauses etc. 
%\enumsentence{Kommer övriga datorer idag som ska vara inte att ha målats?}
%\label{sent:what}
%Sentence (\ref{sent:what})
%PredVP (DetCN Subject (DetSub oovriga_Det) (RelCN (AdvCN (UseN computer_N) (AdvSub today_Adv)) (UseRCl (TTAnt TFut ASimul) PPos (RelVP IdRP UseCopula)))) (PassVP (SlashV2A paint_V2A (ComparA Object red_A (UsePN Object johan_PN))))
%The use of the NONEXIST pattern, as described in \ref{sec:Formal}. When using
%the grammar for generation, an extra

@@

We have developed the main components
for a deep Swedish parser; an extended grammar and lexicon and material for
evaluation and disambiguation. %To abstract:  by combining resources.
We started from a well-defined system for describing grammars as we built
the grammar from the GF resource grammar.

@@

Our final goal is to be able to parse unrestricted text, but considering that
no syntactic theory is yet capable
of wholly covering a natural language, we are satisfied with %theconsider our grammar 
%when used without a robust layer 
the grammar implementation of an important fragment of Swedish.
We have therefore focused on constructions that are frequent in Talbanken, 
and following methods for making the parser robust as described in section \ref{sec:future}
we still believe that it is possible to achieve the goal. 
\\

%TODO more here
Being a natural language processing-project, it will probably never be
entirely complete. Even so, we believe  
%Even though the grammar will not be totally covering, we belive
that the work done is both interesting and useful.
All parts of the project are open-source and may thus be used in other applications.
The grammar may be beneficial also when working with controlled languages,
as it increases the coverage of the Swedish resource grammar.

>>>>

\abstract{
The thesis describes the work formalising an important fragment of Swedish
in the GF formalism and to compare and evaluate the possibilities of using GF
for modelling large-scale grammars. Grammatical constructions specific to and
frequent in Swedish have been investigated and formalised in the functional
programming language.
%~~achieve a covering Swedish~~ grammar implemented
%TODO first sentence should be more, no understatements!
%in GF -- to model natural language by functional programming. 
The idea has been to combine existing language technology resources with
new techniques, and the aim to obtain a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
\\We present the methods and technologies used and further inspects the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and a Swedish grammar implementation. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
extends it to cover constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case for dealing with reflexive pronouns.
%the linguistic phenomena

@@

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed and our
goal is
to create an open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules describing morphology and syntax.
%TODO verify this, check all occurences of grammar
From now on, this is what we mean by grammar, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get %and thereby get
%The usage of GF allows us to start from
a well-defined system for describing language, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

\newpage % until it looks nicer

@@

\section{Outline}
The thesis is divided into 6 chapters. We start by giving some background
information to areas relevant to the project. %Chapter \ref{sec:background} 
Grammatical Framework is presented in section \ref{sec:gf} while a more profound
description of the Swedish resource grammar is given in section \ref{sec:swegf}.
Section \ref{sec:swedish} gives an introduction to the Swedish language, and
brief presentations of Saldo and Talbanken are found in section \ref{sec:saldo} and
\ref{sec:talbanken} respectively.
A summary of related work can be found in section \ref{sec:related}.

@@

In a GF lexicon this verb is given the
category \verb-V3-, %classifing it as 
a \textit{three-place verb}, taking two objects.
The notion \verb-V3- is motivated by the formal translation: \verb-bring(I,it,him)-.\\
%these will be encoded in different ways.
Particles are given in the lexicon, as well as the prepositions that are
chosen by the verb.
The entry for `\emph{ta med}', as used in sentence (\ref{sent:tamed}), is
described as follows:
%, assuming that we already have an constant \verb-take_V- which produces the correct conjugation for ``ta":
\begin{verbatim}
ta_med_V3 = dirV3 (reflV (partV (take_V "med"))) (mkPrep "till") ;
\end{verbatim}
%The category \verb-V3- tells us that the verb takes two objects, % (``den" and ``honom").
The function \verb-dirV3- creates a three-place verb, where the first object is direct and
the second is to be used with the preposition given as the last argument: `\emph{till}'.
\verb-reflV- shows that the verb always is used with a reflexive
pronoun and \verb-partV- gives the particle \emph{`med'}.
The fact that the chosen prepositions is attached to the verb in the lexicon
causes the parse tree visualization algorithm to group them together.
This is also the case for particles, cf. parse tree \ref{pic:tittapart}
and \ref{pic:tittapa}.
As already stated, the visualized parse trees is not a complete representation,
even if the verb phrases in the the visualizations look the same, the two cases
are treated and represented differently internally. The fronting of the
preposition, as in of sentence (\ref{sent:tittapa}a)., is accepted but
fronting of particles, as in \ref{sent:tittapa}b., is not.
%TODO use png to get correct size
\eenumsentence{
\item[a]\shortex{4}
{På &pojken& tittar& du.}
{\emph{on} &\emph{the boy}& \emph{look}& \emph{you}}
{`You look at the boy.'}
\item[b]\shortexnt{6}
{*På& när& du& springer& tittar& jag}
{\emph{~on}& \emph{when}& \emph{you}& \emph{run}& \emph{watch}& \emph{I}}
\label{sent:tittapa}}
\begin{figure}[h]
\subfloat[Verb with a chosen preposition]{\label{pic:tittapa}\includegraphics[height=80mm]{tittapa.jpg}}
\hspace{10mm}
\subfloat[Verb with particle]{\label{pic:tittapart}\includegraphics[height=80mm]{tittapart.pdf}}
\caption{The visualized parse trees do not show the internal difference of chosen prepositions and particles}
\label{fig:translationtrees}
\end{figure}

@@

In addition to the core resource grammars, which is shared with the other languages implemented in the
library, 
%In addition to the basic resource grammar,
 language specific constructions may bee added
%>> core is shared but particular lang. has particular features.
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
Among those were functions for topicalisation: %$ although not all
%of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
%For example, \verb|FocObj| fronted the object as in sentence (\ref{gfSwe:apple}).
\enumsentence{\shortex{6}
{Det där & äpplet & vill & jag & inte & ha}
{that & apple & want & I & not &have}
{`I don't want that apple'}}\label{gfSwe:apple}
and for preposition stranding:
\enumsentence{
~~Stranded preposition \hspace{45mm} cf.\\
\begin{tabular}{ll}
{\begin{tabular}{llllll}
Vem& måste& jag& akta& mig &\textbf{för}? \\
\emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me} &\textbf{\emph{for}}? \\
\end{tabular}}&
{\begin{tabular}{llllll}
\textbf{För}& vem& måste& jag& akta& mig?\\
\emph{\textbf}{for}& \emph{who}& \emph{must}& \emph{I}& \emph{watch out}& \emph{me}?\\
\end{tabular}}\\
~~`Who do I need to watch out \textbf{for}?' &~ `\textbf{For} whom do I need to watch out?'\\
\end{tabular}}

@@

\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw {\sc self's} children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
{\sc self's} gloves, he had forgotten on the train.}
%\enumsentence{Han ger sina pengar till sina syskon.\\
%He gives {\sc self's} money to {\sc self's} children.}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all {\sc self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
a. & Han är längre än sin kompis. & & b. &Han är här oftare än sin kompis.\\
   & He is taller than {\sc self's} friend.
   & & & He is here more often than {\sc self's} friend.
\end{tabular}}\label{ex:ReflAP} \label{ex:ReflAdv}
\eenumsentence{%\begin{tabular}[t]{@{}*{5}{l@{\ }}}
\item Hon tyckte om skolan och alla sina elever.\\
     She liked the school and all {\sc self's} students.
\item Han såg sina få böcker och sin penna.\\
 He saw {\sc self's} few books and {\sc self's} pencil.
}\label{ex:ReflConj} \label{ex:ReflConj2}
Reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in %, it is hence not
%allowed as subject. The antecedent must not be out of scope for the closest finite verb,
sentence (\ref{sent:sesig}) and (\ref{ex:ReflSubj}).
The third person reflexives (`sig',`sin') requires a third person antecedent (see \ref{ex:refljag}).
Furthermore, the antecedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).
The grammar should not accept any of these sentences:
\enumsentence{
*Sina vantar var kvar på tåget.\\
{\sc self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Han och sin kompis läser en bok.\\
He and {\sc self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give {\sc self's} money to {\sc self's} siblings. }\label{ex:refljag}
\enumsentence{*Han vill att jag ser på sig.\\
He wants me to look at {\sc self}.}\label{sent:sesig}

@@

Apart for this, only corrections have been done, exemplified in
this sentence.
\enumsentence{
Hon sover \textbf{som} är bra ~ \textrightarrow ~ Hon sover, \textbf{vilket} är bra\\
``She sleeps, which is good"}

@@

\chapter{Discussion}
\label{sec:results}
\section{Evaluation}
%TODO
%>>evolution diagram?
%also removed unnecessary ambigutiies.
%>> how to handle ambigs: know them and controll them. using statistics or ontolgies.
%>> that is: disambiguitation by context or by meaning.
The project has resulted in
\begin{itemize}
\item a large-scale GF lexicon and a program to redo the importation when needed
\item a comparison between GF and another annotation
\item a deeper testing of the Swedish resource grammar and an estimation
of how well GF can be used to describe larger parts of a language
\item a study of how dependent types can be used in the resource grammars
\end{itemize}

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which is fast even with an extensive lexicon.
%TODO make nice!
Although it cannot not been verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{I'm getting hungry}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{I'm getting hungry}}\label{sent:complVV1}
Since there is no information about which verbs that allows the infinitive mark
to be left out, it will also allow the more questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & låter bli & \textbf{att}& titta} 
{I & let & be &\textbf{to} &watch}
{I don't watch}
\item \shortexnt{4}
{*Jag & låter & bli & titta} 
{I & leave &be & watch}}\label{sent:complVV2}
%ComplBareVV, SupCl
For random generation, the sentences generated may be next to impossible
to interpret without looking at the abstract tree, since they may contain
too many attributes, relative clauses etc. 
%\enumsentence{Kommer övriga datorer idag som ska vara inte att ha målats?}
%\label{sent:what}
%Sentence (\ref{sent:what})
%PredVP (DetCN Subject (DetSub oovriga_Det) (RelCN (AdvCN (UseN computer_N) (AdvSub today_Adv)) (UseRCl (TTAnt TFut ASimul) PPos (RelVP IdRP UseCopula)))) (PassVP (SlashV2A paint_V2A (ComparA Object red_A (UsePN Object johan_PN))))
%The use of the NONEXIST pattern, as described in \ref{sec:Formal}. When using
%the grammar for generation, an extra

@@

We have developed the main components
for a deep Swedish parser; an extended grammar and lexicon and material for
evaluation and disambiguation. %To abstract:  by combining resources.
By starting from the GF resource grammar, we got a well-defined system for
describing grammars. 
%as we built the grammar from .

@@

Our final goal is to be able to parse unrestricted text, but considering that
no syntactic theory is yet capable
of wholly covering a natural language, we are satisfied with %theconsider our grammar 
%when used without a robust layer 
the grammar implementation of an important fragment of Swedish and have
focused on constructions that are frequent in Talbanken. 
Being a natural language processing-project, it will probably never be
entirely complete, but by combining the rule-based parser with statistical
techniques, 
%By following the methods for making the parser robust 
as described in section \ref{sec:future},
we still believe that it is possible to achieve the goal. 
\\

%TODO more here
%. Even so, we believe  
%Even though the grammar will not be totally covering, we belive
%that the work done is both interesting and useful.
All parts of the project are open-source and may thus be used in other applications.
The grammar may be beneficial also when working with controlled languages,
as it increases the coverage of the Swedish resource grammar.
